---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)
```

# studCRIRT

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/studCRIRT)](https://CRAN.R-project.org/package=studCRIRT)
<!-- badges: end -->

The goal of studCRIRT is ...

## Installation

You can install the development version of studCRIRT with:

``` r
# install.packages("devtools")
devtools::install_github("giuseppealfonzetti/studCRIRT")
```

## Evaluate the integrand

At first, we need to specify the setting dimensions

```{r}
library(studCRIRT)
## change the seed to try different draws
seed <- 123

## Setup dimensions
external_covariates <- 1L     # number of external covariates
years_before <- 6L            # number of possible years without graduation risk
years_after <- 2L             # number of possible years  with graduation risk
grades <- 4L                  # number of grades
exams <- 10L                  # number of exams
```

Now we can generate a random vector of paramaters
```{r}

# dimension of the parameter vector related to the competing risk model
dim_cr <- 3*(external_covariates+2) + 2*(years_before) + years_after 

# dimension of the parameter vector related to the irt model
dim_irt <- 3*exams + exams*grades

# dimension of the parameter vector related to latent variables
dim_lat <- 2

# draw a random parameter vector of the suitable length
set.seed(seed)
theta <- rnorm(dim_cr+dim_irt+dim_lat)

# To inspect the parameters we can convert theta to a named list
params_list <- paramsVec2list(THETA = theta,
                              DIM_EXT = external_covariates,
                              NYB = years_before,
                              NYA = years_after,
                              N_GRADES = grades,
                              N_EXAMS = exams)

# here you can manually change parameters (chek the list names)
# I do that just to simulate reasonable data
params_list[['LAT']][['Corr']] <- .5
params_list[['LAT']][['Speed_variability']] <- 1.5
params_list[['IRT']][['Exams_average_time']] <- sapply(sort(rep(1:5, (round(exams/6)+1)))[1:exams], function(x) log(x*180))
params_list[['IRT']][['Exams_variability_time']] <- runif(exams, .5,2)
params_list$IRT$Exams_slopes <- runif(exams, 0.5,1)
params_list$CR$Graduation$Slope_ability <- 2
params_list$CR$Graduation$Slope_speed <- 2

# and project them back on a new theta
theta <- paramsList2vec(PARAMS_LIST = params_list,
                        DIM_EXT = external_covariates,
                        NYB = years_before,
                        NYA = years_after,
                        N_GRADES = grades,
                        N_EXAMS = exams)
```

Generate a the grades and times of a random student for the given parameters
```{r}

## read latent parameters from list
rho <- params_list[['LAT']][['Corr']]
sig <- params_list[['LAT']][['Speed_variability']]

## latent covariance matrix
S <- matrix(c(1,rho*sig,rho*sig,sig^2), 2, 2)

## draw ability, speed and external covariates value 
set.seed(seed+2)
lat <- mvtnorm::rmvnorm(1, sigma = S) # lat[1] = ability, lat[2] = speed
# lat <- matrix(c(.5,.5), ncol = 2)
x <- rnorm(external_covariates)       # external covariates

## draw grades and times for all exams
grades_vec <- rngGrades(
  THETA_IRT = theta[(dim_cr+1):(dim_cr+dim_irt)],
  N_GRADES = grades,
  N_EXAMS = exams,
  ABILITY = lat[1],
  SEED = seed
)$grades
times_vec <- rngTimes(
  THETA_IRT = theta[(dim_cr+1):(dim_cr+dim_irt)],
  N_GRADES = grades,
  N_EXAMS = exams,
  SPEED = lat[2],
  SEED = seed
)

## Read the year of the last exam
yle <- round(max(times_vec)/365,0)
yle
```

Finally, we can plot the integrand function for each outcome and year going from `1` to `yle+1`. Red dots represent the true latent space coordinates. Black dots the ones estimated via `optim`. Grey windows are related to impossible outcomes.
```{r integrandVals}
library(tidyr)
library(dplyr)
library(ggplot2)
library(purrr)

# grid for ability and speed divided by outcome and year
dt <- expand_grid(
  ability = seq(-5,5, by=.25),
  speed = seq(-5,5, by=.25),
  outcome = c(0,1,2,3),
  year = 1:(yle+1)
  )

dtopt <- expand_grid(
  outcome = c(0,1,2,3),
  year = 1:(yle+1)
  ) |> 
  mutate(
    opt = pmap(list(outcome, year),
                   function(x3,x4){

                     # use only exams observed up to `x4`
                     maxTime <- 365*x4
                     obs_vec <- grades_vec>0&times_vec<maxTime
                     Rfun <- function(PAR){
                       -integrand(
                        THETA = theta,
                        EXTCOVARIATES = x,
                        EXAMS_GRADES = grades_vec,
                        EXAMS_DAYS = times_vec,
                        EXAMS_OBSFLAG = obs_vec,
                        OUTCOME = x3,
                        YEAR = x4,
                        N_GRADES = grades,
                        N_EXAMS = exams,
                        NYB = years_before,
                        NYA = years_after,
                        ABILITY = PAR[1],
                        SPEED = PAR[2],
                        YEAR_LAST_EXAM = yle,
                        LOGFLAG = TRUE
                      )
                     }

                     numopt <- try(optim(par = c(0,0), fn = Rfun)$par)
                     numopt

    })
  ) |> 
  mutate(error = map_lgl(opt, ~class(.x)[1]=="try-error")) |> 
  mutate(opt = map2(error, opt, ~if(.x){c(NA,NA)}else{.y})) |> 
  mutate(
    ability = map_dbl(opt, ~.x[1]),
    speed = map_dbl(opt, ~.x[2])
  )

# compute integrand value on the grid
dt_val <- dt |>
  mutate(
    val = pmap_dbl(list(ability, speed, outcome, year),
                   function(x1,x2,x3,x4){
                     
                     # use only exams observed up to `x4`
                     maxTime <- 365*x4
                     obs_vec <- grades_vec>0&times_vec<maxTime
                     
                     integrand(
                        THETA = theta,
                        EXTCOVARIATES = x,
                        EXAMS_GRADES = grades_vec,
                        EXAMS_DAYS = times_vec,
                        EXAMS_OBSFLAG = obs_vec,
                        OUTCOME = x3,
                        YEAR = x4,
                        N_GRADES = grades,
                        N_EXAMS = exams,
                        NYB = years_before,
                        NYA = years_after,
                        ABILITY = x1,
                        SPEED = x2,
                        YEAR_LAST_EXAM = yle,
                        LOGFLAG = TRUE
                      )
      
    })
    
  )

# rearrange and plot
# the integrand value is scaled just for visual purposes to have 
# similar color-scales across facets
dt_val |>
  group_by(year, outcome) |>
  mutate(scaledIntegrand = scale(val),
         outcome = factor(outcome,
                          levels = c(0,1,2,3),
                          labels = c('Enrollment', 'Dropout', 'Transfer', 'Graduation'), 
                          ordered = T)) |>
  ggplot(aes(x = ability, y = speed)) +
  geom_raster(aes(fill = scaledIntegrand)) +
  geom_point(data = as_tibble(lat) |> rename(ability = V1, speed = V2), col = 'red')+
  geom_point(data = dtopt |> mutate(
         outcome = factor(outcome,
                          levels = c(0,1,2,3),
                          labels = c('Enrollment', 'Dropout', 'Transfer', 'Graduation'), 
                          ordered = T)), 
         aes(x = ability, y = speed), col = 'black') +
  facet_grid(year~outcome)
```



